---
title: "tekrarlı ölçümler deneme deseni"
author: "Assoc. Prof. Burcu MESTAV"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: yes
    number_sections: yes
    toc: true
    toc_depth: 2
    highlight: textmate
    html_preview: TRUE
    df_print: kable
    pandoc_args:
    - --number-sections
    - --number-offset=0
fontsize: 11pt
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	dpi=500,
	dev = "jpeg",
	fig.width = 14,
	fig.height = 10
)
```



```{r Library-calls, message=FALSE, warning=FALSE, include=F, eval=T, results="hide"}
library(tidyverse)
library(olsrr)
library(data.table)
library(ggthemr) # harika bir tema
library(kableExtra)
library(knitr)
library(PerformanceAnalytics)
library(GGally)
library(ggpubr)
library(agricolae)
library(MASS)
library(tidyr)
library(broom)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
library(downloadthis)
library(car)
library(downloadthis)
library(readxl)
library(stringr)
library(ggstatsplot)
library(rstatix)
library(nlme)
library(lme4)
```


```{r fonksiyon, include=FALSE}
create_dt <- function(x){
  DT::datatable(x,
                extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                               buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                               lengthMenu = list(c(10,25,50,-1),
                                                 c(10,25,50,"All"))))
}
```

# Giriş

```{r}

#data2 <- read.csv('http://courses.washington.edu/psy524a/datasets/SubjectExerciseANOVAdependent.csv')

#data2.wide <- reshape(data2, idvar ="subject", timevar = "time", direction = "wide")

dat=read.csv("dat.csv")
dat2=read.csv("dat2.csv")

```


```{r}
anova.out.3 <- anova(lmer(weight ~ time + (1|subject),data = data2))
anova.out.3

rm_anova_nlme <- lme(
  weight ~ time,
  random = ~1 | subject,
  data = data2
)

# Print the results
print(summary(rm_anova_nlme))
# Print the results
print(anova(rm_anova_nlme))



# Perform Mixed ANOVA

model = aov(weight ~ time + Error(subject/time), data = data2)

summary(model)
```

## Kalp atış hızı (1RM)

```{r}
library(tibble)

# Create a dataframe with medical application data in wide format
medical_data <- tribble(
  ~hasta_no, ~Önce, ~Sırasında, ~Sonra,
  1, 75, 85, 80,
  2, 80, 90, 82,
  3, 70, 75, 74,
  4, 85, 95, 88,
  5, 78, 88, 81
)
create_dt(medical_data)
```

### PREPARING DATA
Before exploring and modelling ANOVA is necessary to pass the data to long format.

To convert the wide format data to long format, we can use the pivot_longer function from the tidyr package:

```{r}
# Load necessary packages
library(tidyr)

# Convert to long format
medical_data <- pivot_longer(
  data = medical_data,
  cols = c(Önce, Sırasında, Sonra),
  names_to = "Zaman_noktası",
  values_to = "Kalp_Hızı"
)
```

And converting Patient ID to factor (Although for certain packages is not necessary):

```{r}
medical_data$hasta_no = as.factor(medical_data$hasta_no)

medical_data$Zaman_noktası = factor(medical_data$Zaman_noktası,
                                    levels=c("Önce","Sırasında","Sonra"))

# Print the dataframe in long format
print(medical_data)
```

### DATA EXPLORATION
Let’s create some tables to summarize the data

```{r}
table_sum<- medical_data %>%
  group_by(Zaman_noktası) %>%
  get_summary_stats(Kalp_Hızı, type = "full",
  show = c("n","mean","sd","max","min","se","median"))
create_dt(table_sum)
```


```{r}
ggplot(medical_data, aes(Zaman_noktası,Kalp_Hızı))+
  geom_boxplot(aes(col=Zaman_noktası))+
  labs(title="5 Hastanın Ameliyat Öncesi, Sırasında ve Sonrasında Ölçülen Kalp Atış Hızları ", x=" ", y="Kalp Hızı")+
  theme_bw()+
  guides(fill=F) + geom_hline(yintercept=mean(medical_data$Kalp_Hızı),
                              linetype="dotted")+
  theme(text = element_text(family = "Times New Roman", size=11),legend.position = " ", 
        legend.title = element_blank())
  

```

```{r}
# Create an interaction plot
interaction.plot(
  x.factor = medical_data$Zaman_noktası,
  trace.factor = medical_data$hasta_no,
  response = medical_data$Kalp_Hızı,
  xlab="Zaman",
  ylab = "Kalp Atış Hızı",
  trace.label = "Hasta"
)

ggplot(medical_data, aes(Zaman_noktası, Kalp_Hızı, colour=hasta_no)) +
  geom_point(size = 2.5) +
  geom_line(aes(group = hasta_no), linewidth = 1) +
  theme_minimal()
```

```{r}
anova.out.3 <- anova(lmer(Kalp_Hızı ~ Zaman_noktası + (1|hasta_no),data = medical_data))
anova.out.3

rm_anova_nlme <- lme(
  Kalp_Hızı ~ Zaman_noktası,
  random = ~1 | hasta_no,
  data = medical_data
)

# Print the results
print(anova(rm_anova_nlme))
# Print the resultsx

print(anova(rm_anova_nlme))



# Perform Mixed ANOVA

model = aov(Kalp_Hızı ~ Zaman_noktası+Error(hasta_no/Zaman_noktası) , data = medical_data)

summary(model)

# Perform Repeated Measures ANOVA with the rstatix package
rm_anova <- anova_test(
  data = medical_data,
  dv = Kalp_Hızı,
  wid = hasta_no,
  within = Zaman_noktası,
  type = 3  # Specifies repeated measures design
)

# Print the results
rm_anova

pwc <- medical_data %>%
  pairwise_t_test(
    Kalp_Hızı ~ Zaman_noktası, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
create_dt(pwc)


pwc <- pwc %>% add_xy_position(x = "Zaman_noktası")
bxp + 
  stat_pvalue_manual(pwc) +
  labs(
    subtitle = get_test_label(rm_anova, detailed = TRUE),
    caption = get_pwc_label(pwc)
  )
```


## two way on one repeated factor 

```{r}
Input = ("
Katılımcı	Program	Ay	Kalori
1	ProgramA	Ay1	2000
2	ProgramA	Ay1	1900
3	ProgramA	Ay1	2100
4	ProgramA	Ay1	2000
1	ProgramA	Ay2	1978
2	ProgramA	Ay2	1826
3	ProgramA	Ay2	2067
4	ProgramA	Ay2	1981
1	ProgramA	Ay3	1962
2	ProgramA	Ay3	1782
3	ProgramA	Ay3	2065
4	ProgramA	Ay3	1987
1	ProgramA	Ay4	1873
2	ProgramA	Ay4	1718
3	ProgramA	Ay4	2015
4	ProgramA	Ay4	2016
1	ProgramA	Ay5	1782
2	ProgramA	Ay5	1639
3	ProgramA	Ay5	1994
4	ProgramA	Ay5	2010
1	ProgramA	Ay6	1737
2	ProgramA	Ay6	1644
3	ProgramA	Ay6	1919
4	ProgramA	Ay6	1946
1	ProgramB	Ay1	2100
2	ProgramB	Ay1	2000
3	ProgramB	Ay1	2000
4	ProgramB	Ay1	2000
1	ProgramB	Ay2	2004
2	ProgramB	Ay2	2011
3	ProgramB	Ay2	2074
4	ProgramB	Ay2	1970
1	ProgramB	Ay3	2027
2	ProgramB	Ay3	2089
3	ProgramB	Ay3	2141
4	ProgramB	Ay3	1951
1	ProgramB	Ay4	2109
2	ProgramB	Ay4	2124
3	ProgramB	Ay4	2199
4	ProgramB	Ay4	1981
1	ProgramB	Ay5	2197
2	ProgramB	Ay5	2199
3	ProgramB	Ay5	2265
4	ProgramB	Ay5	1987
1	ProgramB	Ay6	2294
2	ProgramB	Ay6	2234
3	ProgramB	Ay6	2254
4	ProgramB	Ay6	1969
1	ProgramC	Ay1	1950
2	ProgramC	Ay1	2000
3	ProgramC	Ay1	2000
4	ProgramC	Ay1	2000
1	ProgramC	Ay2	2007
2	ProgramC	Ay2	2029
3	ProgramC	Ay2	1976
4	ProgramC	Ay2	2020
1	ProgramC	Ay3	1978
2	ProgramC	Ay3	2033
3	ProgramC	Ay3	2025
4	ProgramC	Ay3	2009
1	ProgramC	Ay4	1965
2	ProgramC	Ay4	2050
3	ProgramC	Ay4	2047
4	ProgramC	Ay4	2017
1	ProgramC	Ay5	1984
2	ProgramC	Ay5	2001
3	ProgramC	Ay5	2033
4	ProgramC	Ay5	1989
1	ProgramC	Ay6	2020
2	ProgramC	Ay6	1988
3	ProgramC	Ay6	1984
4	ProgramC	Ay6	2020
")
data = read.table(textConnection(Input),header=TRUE)
data$Program = as.factor(data$Program)
data$Katılımcı = factor(data$Katılımcı,
                         levels=unique(data$Katılımcı))
data$Ay = factor(data$Ay,
                         levels=unique(data$Ay))

summary(data)
```

```{r}
p1box <- ggplot(data=data, aes(x=Ay, y=Kalori, fill=Program)) +
  geom_boxplot()
p2box <- p1box +
  labs(title="Program ve Aya göre Kalori değerleri", x="Ay", y = "Kalori değeri") +
  theme_classic() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.background = element_blank(),
        axis.line.y = element_line(colour="black", linewidth=.7),
        axis.line.x = element_line(colour="black", linewidth=.7),
        plot.title = element_text(hjust=.5)
)+
  scale_fill_manual(values=c('honeydew3','honeydew2', 'honeydew1'))
p2box
```



```{r}
table_sum<- data %>%
  group_by(Program, Ay) %>%
  get_summary_stats(Kalori, type = "full",
  show = c("n","mean","sd","max","min","se","median"))
create_dt(table_sum)

summ <- Rmisc::summarySEwithin(data=data, idvar="Katılımcı",measurevar="Kalori", withinvars=c("Program", "Ay"))
colnames(summ) <- c("Program", "Ay", "N", "mean", "sd", "sem", "CI" )
summ
```

```{r}
interaction.plot(
  x.factor = data$Ay,
  trace.factor = data$Program,
  response = data$Kalori,
  fixed = F,
  xlab="Aylar",
  ylab = "Kalori",
  trace.label = "Beslenme Programı"
)

ggplot(data, aes(Ay, Kalori, colour=Program)) +
  geom_point(size = 1.5) +
  geom_line(aes(group = Katılımcı), linewidth = 1) +
  theme_minimal()
a <- aov_ez("Katılımcı", "Kalori", data, between = "Program", within = "Ay",fun_aggregate = NULL)

afex_plot(a, "Program", "Ay", "Kalori")

```


```{r}
fit1.afex <- aov_car(Kalori ~ (Program*Ay)
                     + Error(Katılımcı/(Program*Ay)),
                     anova_table = list(correction = "none"),
                     return="univariate",
                     data=data)
gt(afex::nice(fit1a.afex, es=c("pes","ges"), correction = "none"))
```

```{r}
res.aov <- data %>% anova_test(dv = Kalori,wid = Katılımcı, within = c(Program,Ay))
```

